A forest is more than its trees: haplotypes and inferred ARGs

### 2024-12-16 paper/writeup.tex
do something about parsimony: maybe closes #16

\paragraph{Ignorance and omission in an ARG}
As motivation, we presented above a ``historical'' view of ARGs --
i.e., that each aspect of an inferred ARG is intended to represent
a portion of some particular historical genome
(for instance, the MRCA of some set of sampled genomes).
Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
that relationships \emph{not} depicted in an ARG are implied to not exist.
An alternative interpretation
of the ARG depicted in \figref[A\&B]{fig:conceptual} would be that we have no information as to 
how node 2 inherited from node 4 on the right-hand interval,
rather than saying that the line of transmission specifically did not pass
through node 3.
The ``simplification'' algorithm~\citep{kelleher2018efficient}
and the Hudson algorithm for coalescent
simulation~\citep{hudson1983properties,kelleher2016efficient}
each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;
so for ARGs produced by these algorithms, the correct interpretation is that
the omission of unary spans reflects a lack of knowledge.
In this paper, we have shown that, for the most part, this missing information can be imputed.

\paragraph{Parsimony}
Much of the early work on ARG inference aimed to extract as much information as possible
out of the small datasets of the time,
and so, roughly speaking, integrated over possible ARGs
with the goal of inferring higher-level parameters:
mostly, scaled mutation rate and recombination rate
\citep[e.g.,][]{hudson1985statistical,griffiths1996ancestral,kuhner2000maximum,stephens2000inference,fearnhead2001estimating}.
However, the space of possible ARGs for a given dataset is extremely large,
and other work aimed to identify the minimum number of recombinations
needed to explain a given dataset under the infinite alleles model of mutation
\citep[e.g.,][]{hein1990reconstructing,myers2003bounds,song2005constructing},
which turns out to be NP-complete \citep{wang2001perfect}.
For instance, \citet{minichiello2006mapping}
used a heuristic algorithm to produce ``plausible'' ARGs
(i.e., those that explained the data with few mutations and recombinations),
and searched for associations with traits in the resulting ensemble of ARGs.
Our approach for extending haplotypes follows the same logic,
that an ARG with fewer recombination events is more parsimonious,
and thus more likely.
For this reason, it will occasionally be wrong even if the trees are correct,
although in practice this source of error is likely much smaller than error in tree inference itself.

>>>>

\paragraph{Ignorance and omission in an ARG}
As motivation, we presented above a ``historical'' view of ARGs --
i.e., that each aspect of an inferred ARG is intended to represent
a portion of some particular historical genome
(for instance, the MRCA of some set of sampled genomes).
Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
that relationships \emph{not} depicted in an ARG are implied to not exist.
As discussed in \citet{wong2024general},
an alternative interpretation
of the ARG depicted in \figref[A\&B]{fig:conceptual} would be that we have no information as to 
how node 2 inherited from node 4 on the right-hand interval,
rather than saying that the line of transmission specifically did not pass
through node 3.
The ``simplification'' algorithm~\citep{kelleher2018efficient}
and the Hudson algorithm for coalescent
simulation~\citep{hudson1983properties,kelleher2016efficient}
each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;
so for ARGs produced by these algorithms, the correct interpretation is that
the omission of unary spans reflects a lack of knowledge.
In this paper, we have shown that, for the most part, this missing information can be imputed.

\paragraph{Parsimony}
Much of the early work on ARG inference aimed to extract as much information as possible
out of the small datasets of the time,
and so, roughly speaking, integrated over possible ARGs
with the goal of inferring higher-level parameters:
mostly, scaled mutation rate and recombination rate
(for instance, \citet{hudson1985statistical,griffiths1996ancestral,kuhner2000maximum,stephens2000inference,fearnhead2001estimating}).
However, the space of possible ARGs for a given dataset is extremely large,
and other work aimed to identify the minimum number of recombinations
needed to explain a given dataset under the infinite alleles model of mutation
\citep[e.g.,][]{hein1990reconstructing,myers2003bounds,song2005constructing},
which turns out to be NP-complete \citep{wang2001perfect}.
For instance, \citet{minichiello2006mapping}
used a heuristic algorithm to produce ``plausible'' ARGs
(i.e., those that explained the data with few mutations and recombinations),
and searched for associations with traits in the resulting ensemble of ARGs.
(See \citet{wong2024general} for more historical discussion.)
Our approach for extending haplotypes follows the same logic,
that an ARG with fewer recombination events is more parsimonious,
and thus more likely.
For this reason, it will occasionally be wrong even if the trees are correct,
although in practice this source of error is likely much smaller than error in tree inference itself.
